---
title: "Vacant Building Intro"
author: "TJ Butler"
format: html
editor: visual
execute:
  cache: true
---

```{r}
#| label: setup
#| include: false

if(!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse", "sf", "mapview", "gridExtra", "survey", "devtools",
               "ggmosaic", "basemaps", "tidycensus", "tmap", "basemaps",
               "maptiles", "tidygeocoder", "openxlsx2")

# Get basemap using maptiles
osm_tiles <- get_tiles(vbnCityData, provider = "OpenStreetMap",
                       zoom = 12, crop = T, cachedir = "osm_tiles")
```

```{r}
#| label: setup2
#| include: false
vbnCityData <- 
st_read("https://egisdata.baltimorecity.gov/egis/rest/services/Housing/DHCD_Open_Baltimore_Datasets/FeatureServer/1/query?outFields=*&where=1%3D1&f=geojson")

csaBoundaries <- 
st_read("spatial_data/Total_Population/Total_Population.shp")

csa_pop_count <- csaBoundaries %>% 
  as.data.frame(.) %>% 
  select(CSA2010:tpop20)

tract_to_csa <- 
  st_read("https://services1.arcgis.com/mVFRs7NF4iFitgbY/arcgis/rest/services/Tract2020_to_CSA2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
  as.data.frame(.) %>% select(-geometry) %>% 
  filter(Community_Statistical_Area_2020 != "Unassigned -- Jail")

load("BAS_data/baltimore-area-survey-2023.Rdata")
load("BAS_data/baltimore-area-survey-2024.Rdata")
load("BAS_data/baltimore-area-survey-2025.Rdata")

#Filter down to city for BAS24 and 25
bas24 <- bas24 %>% 
  filter(bas24_svy_cty == "City")

bas25 <- bas25 %>% 
  filter(bas25_svy_cty == "City")

# Set up Census API key - already set up
#census_api_key("b2cedbd221d7cb2c24f7f90d54ab3e3d79bdf929", install = T)

# Get ACS data for num buildings by census tract
acs_ct <- get_acs(geography = "tract", state = "MD", county = "Baltimore city",
                   variables = c(housing_units = "B25001_001"), 
                   year = 2024, geometry = T, survey = "acs5")

acs_ct <- st_transform(acs_ct, 4326)

st_crs(acs_ct)

acs_vacant_per_1k_blds <- acs_ct %>% 
  mutate(vacant_num = lengths(st_intersects(acs_ct, vbnCityData)),
         vacant_per_1k_bldg = round(vacant_num / estimate * 1000))

# Mapping census tract to CSA
acs_to_csa <- left_join(acs_vacant_per_1k_blds, tract_to_csa, 
                        join_by(GEOID == GEOID_Tract_2020)) %>% 
  filter(!is.na(Tract_2020)) %>% 
  group_by(Community_Statistical_Area_2020) %>% 
  summarize(num_blds = sum(estimate),
            vacant_num = sum(vacant_num)) %>% 
  mutate(vac_bldg_per_1k = round(vacant_num / num_blds * 1000),
         quartile = case_when(
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.25) ~ 1,
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.5) ~ 2,
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.75) ~ 3,
           T ~ 4
         ),
         quintile = case_when(
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.2) ~ 1,
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.4) ~ 2,
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.6) ~ 3,
           vac_bldg_per_1k <= quantile(vac_bldg_per_1k, 0.8) ~ 4,
           T ~ 5
         ))

# Num Responses by CSA as tibble
n_responses_bas24 <- bas24 %>% group_by(bas24_svy_csaid) %>% count() %>% 
  rename(n_bas24 = n)
n_responses_bas25 <- bas25 %>% group_by(bas25_svy_csaid) %>% count() %>% 
  rename(n_bas25 = n)

n_responses <- full_join(n_responses_bas24, n_responses_bas25, 
                         join_by(bas24_svy_csaid == bas25_svy_csaid))

# sum(n_responses$n_bas24)
# sum(n_responses$n_bas25)

# Merge to acs_to_csa object to get spatial component
acs_to_csa2 <- full_join(acs_to_csa, n_responses, 
                         join_by(Community_Statistical_Area_2020 == 
                                  bas24_svy_csaid)) %>% 
  full_join(., csa_pop_count, 
            join_by(Community_Statistical_Area_2020 == CSA2010))

acs_to_csa_final <- acs_to_csa2 %>% 
  mutate(n_per_1k_bas24 = round(n_bas24/tpop20*1000,2),
         n_per_1k_bas25 = round(n_bas25/tpop20*1000,2))
```

## Baseplot of VBN's as points

```{r}
#| label: Basemap Practice
#| eval: false


# NOTES: Dot density - 1 dot = X>1 buildings for example
tm_shape(osm_tiles) +
  tm_title("Vacant Building Notices overlaying CSAs") +
  tm_rgb() +
  tm_shape(csaBoundaries) +
  tm_polygons(fill_alpha = 0.6) +
  tm_shape(vbnCityData) +
  tm_dots(size = 0.005, fill = "darkred", fill_alpha = 0.05)
```

## VBN's per 1000 buildings
### Using ACS 5-year 2020-2024 data for number of buildings by Census tract
### Using VBN notices from Baltimore City Open Data

```{r}
#| label: Mapping VBNs per 1000 buildings

# Used ACS For # buildings, VBN from open baltimore for vacant building count

tm_shape(osm_tiles) +
  tm_rgb() +
  tm_shape(acs_vacant_per_1k_blds) +
  tm_polygons(fill = "vacant_per_1k_bldg",
              fill.scale = tm_scale(),
              fill_alpha = 0.5)

tm_shape(acs_to_csa) +
  tm_polygons(fill_alpha = 0.6,
              fill = "vac_bldg_per_1k")
```

## BAS Sentiment Histogram

```{r}
#| label: BAS Sentiment Histogram

# Percentage of response option
bas24_vac_lot <- bas24 %>%
  count(bas24_nhd_pdvlot) %>% 
  mutate(Percent = round(n / nrow(bas24) * 100, 1))

bas24_vac_bldg <- bas24 %>% 
  count(bas24_nhd_pdvbldg) %>% 
  mutate(Percent = round(n / nrow(bas24) * 100, 1))

bas25_vac_bldg <- bas25 %>% 
  count(bas25_nhd_pbvbld) %>% 
  mutate(Percent = round(n / nrow(bas25) * 100, 1))

vac_lot24 <- ggplot(data = bas24_vac_lot, 
                    mapping = aes(x = bas24_nhd_pdvlot, y = Percent)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS24 Vac Lot") + ylim(c(0,60))

vac_bldg24 <- ggplot(data = bas24_vac_bldg, 
                     mapping = aes(x = bas24_nhd_pdvbldg, 
                                   y = Percent)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS24 Vac Bldg") + ylim(c(0,60))

vac_bldg25 <- ggplot(data = bas25_vac_bldg, 
                     mapping = aes(x = bas25_nhd_pbvbld, y = Percent)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS25 Vac Bldg") + ylim(c(0,60))

gridExtra::grid.arrange(vac_lot24, vac_bldg24, vac_bldg25, ncol = 3)
```


## BAS 24 and 25 Total Responses per CSA
NOTE: Response rate by CSA? Do we have that info?
```{r}
#| label: Number and Rate Responses by CSA

# QUESTION: Does 21CCI have number of responses mailed to each CSA, to review 
# response rate by CSA??? 

tm_shape(osm_tiles) +
  tm_rgb() +
  tm_title("Num Responses per CSA BAS24") +
  tm_shape(acs_to_csa_final) +
  tm_polygons(fill = "n_bas24",
              fill.scale = tm_scale(),
              fill_alpha = 0.75)

tm_shape(osm_tiles) +
  tm_rgb() +
  tm_title("Rate/1000 Responses per CSA BAS24") +
  tm_shape(acs_to_csa_final) +
  tm_polygons(fill = "n_per_1k_bas24",
              fill_alpha = 0.75)

tm_shape(osm_tiles) +
  tm_rgb() +
  tm_title("Num Responses per CSA BAS25") +
  tm_shape(acs_to_csa_final) +
  tm_polygons(fill = "n_bas25",
              fill.scale = tm_scale(),
              fill_alpha = 0.75)

tm_shape(osm_tiles) +
  tm_rgb() +
  tm_title("Rate/1000 Responses per CSA BAS25") +
  tm_shape(acs_to_csa_final) +
  tm_polygons(fill = "n_per_1k_bas25",
              fill_alpha = 0.75)
```


## Distribution of VBN's per 1000 buildings stratified by Vacant Building Sentiment Response: 24 25

```{r}
#| label: VBN by vacant building sentiment cross-tab

# Merge vacant buildings per 1000 TO bas data: Census Tract
bas24_acs_joined <- 
  left_join(bas24, acs_vacant_per_1k_blds, join_by(bas24_svy_geoid == GEOID))



ggplot(bas24_acs_joined) +
  geom_histogram(aes(x = vacant_per_1k_bldg))

ggplot(bas24_acs_joined) +
  geom_histogram(aes(x = vacant_per_1k_bldg)) +
  facet_wrap(~bas24_nhd_pdvbldg)


# Vacant buildings per 1000: CSA level
bas24_csa_joined <- 
  left_join(bas24, acs_to_csa, 
            join_by(bas24_svy_csaid == Community_Statistical_Area_2020))

ggplot(bas24_csa_joined) +
  geom_histogram(aes(x = vac_bldg_per_1k))

ggplot(bas24_csa_joined) +
  geom_histogram(aes(x = vac_bldg_per_1k)) +
  facet_wrap(~bas24_nhd_pdvbldg)

# BAS 25
bas25_csa_joined <-
  left_join(bas25, acs_to_csa,
            join_by(bas25_svy_csaid == Community_Statistical_Area_2020))

ggplot(bas25_csa_joined) +
  geom_histogram(aes(x = vac_bldg_per_1k))

ggplot(bas25_csa_joined) +
  geom_histogram(aes(x = vac_bldg_per_1k)) +
  facet_wrap(~bas25_nhd_pbvbld)

# Discordant responses - high vacants/not a problem, low vacants/major problem
# NOTES: Take CSA's in top quartile of # vbn's, look at distribution of responses
# Code response values as 1 through 4 or 5
# Plot correlation of vbn quartile and response code 1-5, select discordant
# section of graph
# Even if no mismatch, how do VBN counts affect things like safety, trust in gov,
# other survey data questions

# HIDE FOR NOW: Discordant CSAs
# bas24_low_vac_big_prob <- bas24_csa_joined %>% 
#   filter(bas24_nhd_pdvbldg == "A big problem" & quartile == 1) %>% 
#   st_as_sf(.)
# 
# tm_shape(osm_tiles) +
#   tm_rgb() +
#   tm_shape(bas24_low_vac_big_prob) +
#   tm_polygons()
# 
# bas24_hi_vac_lo_prob <- bas24_csa_joined %>% 
#   filter(bas24_nhd_pdvbldg == "Not a problem at all" & quartile == 4) %>% 
#   st_as_sf(.)
# 
# tm_shape(osm_tiles) +
#   tm_rgb() +
#   tm_shape(bas24_hi_vac_lo_prob) +
#   tm_polygons()
```

## Discordant Values Graph
```{r}
# Helpful to label # vacants in each quintile??
bldg_sent24_vs_quartile_vbn <- bas24_csa_joined %>%
  filter(!str_detect(bas24_nhd_pdvbldg, "Missing")) %>% 
  group_by(bas24_nhd_pdvbldg, quartile) %>% 
  count()

ggplot(bldg_sent24_vs_quartile_vbn, 
       aes(x = bas24_nhd_pdvbldg, y = quartile, fill = n)) +
  geom_tile()

bldg_sent25_vs_quintile_vbn <- bas25_csa_joined %>% 
  filter(!str_detect(bas25_nhd_pbvbld, "Missing")) %>% 
  group_by(bas25_nhd_pbvbld, quintile) %>% 
  count()

ggplot(bldg_sent25_vs_quintile_vbn, 
       aes(x = bas25_nhd_pbvbld, y = quintile, fill = n)) +
  geom_tile()
```

## Correlation between VBN Metrics and Different Questions
```{r}
#| label: Reviewing correlation between VBN Count/Quartile/Quintile and Neighborhood Safety and Cohesion questions

# Correlation with VBN quartile/quintile and VBN sentiment
bas24_vbn_not_missing <- bas24_csa_joined %>% 
  filter(!str_detect(bas24_nhd_pdvbldg,"Missing"))

cor(bas24_vbn_not_missing$vac_bldg_per_1k,
    as.numeric(bas24_vbn_not_missing$bas24_nhd_pdvbldg))
  
# General solution: limit dataset to non-missing values, do correlation, save
# value with column name - loop over all variables
# First var: bas24_nhd_sat

# BAS 24
i <- 3
cur_data_missing_removed <- bas24_csa_joined %>% 
  filter(bas24_csa_joined %>% pull(i) != "Missing: Item non-response")

vac_per_1k_cor <- 
  cor(cur_data_missing_removed$vac_bldg_per_1k,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_quartile_cor <- 
  cor(cur_data_missing_removed$quartile,
      as.numeric(cur_data_missing_removed %>% pull(i)))
  
vac_quintile_cor <-
  cor(cur_data_missing_removed$quintile,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_sentiment_cor <-
  cor(as.numeric(cur_data_missing_removed$bas24_nhd_pdvbldg),
      as.numeric(cur_data_missing_removed %>% pull(i)))

cor_vars_bas24 <- tibble(var_name = colnames(bas24_csa_joined)[i],
                   num_responses = dim(cur_data_missing_removed)[1],
                   vacants_per_1k_bldgs = vac_per_1k_cor,
                   quartile_vacants = vac_quartile_cor,
                   quintile_vacants = vac_quintile_cor,
                   vacant_sentiment = vac_sentiment_cor)

for(i in 4:143){
cur_data_missing_removed <- bas24_csa_joined %>% 
  filter(bas24_csa_joined %>% pull(i) != "Missing: Item non-response")

vac_per_1k_cor <- 
  cor(cur_data_missing_removed$vac_bldg_per_1k,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_quartile_cor <- 
  cor(cur_data_missing_removed$quartile,
      as.numeric(cur_data_missing_removed %>% pull(i)))
  
vac_quintile_cor <-
  cor(cur_data_missing_removed$quintile,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_sentiment_cor <-
  cor(as.numeric(cur_data_missing_removed$bas24_nhd_pdvbldg),
      as.numeric(cur_data_missing_removed %>% pull(i)))

cor_vars_bas24 <- add_row(cor_vars_bas24, 
                    var_name = colnames(bas24_csa_joined)[i],
                    num_responses = dim(cur_data_missing_removed)[1],
                    vacants_per_1k_bldgs = vac_per_1k_cor,
                    quartile_vacants = vac_quartile_cor,
                    quintile_vacants = vac_quintile_cor,
                    vacant_sentiment = vac_sentiment_cor)
}

cor_vars_bas24 <- arrange(cor_vars_bas24, desc(abs(vacants_per_1k_bldgs)))

# BAS 25

i <- 3
cur_data_missing_removed <- bas25_csa_joined %>% 
  filter(!is.na(bas25_csa_joined %>% pull(i)) & 
           bas25_csa_joined %>% pull(i) != "Missing: Item non-response")

vac_per_1k_cor <- 
  cor(cur_data_missing_removed$vac_bldg_per_1k,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_quartile_cor <- 
  cor(cur_data_missing_removed$quartile,
      as.numeric(cur_data_missing_removed %>% pull(i)))
  
vac_quintile_cor <-
  cor(cur_data_missing_removed$quintile,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_sentiment_cor <-
  cor(as.numeric(cur_data_missing_removed$bas25_nhd_pbvbld),
      as.numeric(cur_data_missing_removed %>% pull(i)))

cor_vars_bas25 <- tibble(var_name = colnames(bas25_csa_joined)[i],
                   num_responses = dim(cur_data_missing_removed)[1],
                   vacants_per_1k_bldgs = vac_per_1k_cor,
                   quartile_vacants = vac_quartile_cor,
                   quintile_vacants = vac_quintile_cor,
                   vacant_sentiment = vac_sentiment_cor)

for(i in 4:189){
cur_data_missing_removed <- bas25_csa_joined %>% 
  filter(bas25_csa_joined %>% pull(i) != "Missing: Item non-response")

vac_per_1k_cor <- 
  cor(cur_data_missing_removed$vac_bldg_per_1k,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_quartile_cor <- 
  cor(cur_data_missing_removed$quartile,
      as.numeric(cur_data_missing_removed %>% pull(i)))
  
vac_quintile_cor <-
  cor(cur_data_missing_removed$quintile,
      as.numeric(cur_data_missing_removed %>% pull(i)))

vac_sentiment_cor <-
  cor(as.numeric(cur_data_missing_removed$bas25_nhd_pbvbld),
      as.numeric(cur_data_missing_removed %>% pull(i)))

cor_vars_bas25 <- add_row(cor_vars_bas25, 
                    var_name = colnames(bas25_csa_joined)[i],
                    num_responses = dim(cur_data_missing_removed)[1],
                    vacants_per_1k_bldgs = vac_per_1k_cor,
                    quartile_vacants = vac_quartile_cor,
                    quintile_vacants = vac_quintile_cor,
                    vacant_sentiment = vac_sentiment_cor)
}

cor_vars_bas25 <- arrange(cor_vars_bas25, desc(abs(vacants_per_1k_bldgs)))


# write_xlsx(list(bas24 = cor_vars_bas24, bas25 = cor_vars_bas25),
#            "correlation_matrix_bas24_25.xlsx")
```


## Mapping the different responses: where do people live?

```{r}
#| label: Maps of response categories for BAS24

# NOTES: Merge in VBN dot density map to bring in actual vacant lot viz data

# Getting counts of Census Tract by response category
bas24_csa_counts <- bas24_csa_joined %>% 
  group_by(bas24_nhd_pdvbldg, bas24_svy_csaid, geometry) %>% 
  summarize(count = n()) %>% 
  st_sf(.)


tm_shape(osm_tiles) +
  tm_rgb() +
  tm_shape(bas24_csa_counts) +
  tm_polygons(fill = "count") +
  tm_facets_wrap(by = "bas24_nhd_pdvbldg") +
  tm_shape(vbnCityData) +
  tm_dots(size = 0.005, fill = "darkred", fill_alpha = 0.05)
```



## Demographic Cross Tabs

```{r}
#| label: Demographic cross tab

vac_lot_race24 <- ggplot(data = bas24, mapping = aes(x = bas24_nhd_pdvlot)) + 
  geom_bar() + theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS 24 Vac Lot Sentiment") + facet_wrap(~bas24_dem_raceeth4)

vac_lot_race24

vac_bldg_race24 <- ggplot(data = bas24, mapping = aes(x = bas24_nhd_pdvbldg)) + 
  geom_bar() + theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS 24 Vac Bldg Sentiment") + facet_wrap(~bas24_dem_raceeth4)

vac_bldg_race25 <- ggplot(data = bas25, mapping = aes(x = bas25_nhd_pbvbld)) + 
  geom_bar() + theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS 25 Vac Bldg Sentiment") + facet_wrap(~bas25_dem_raceeth4)

vac_bldg_race25

vac_lot_sex24 <- ggplot(data = bas24, mapping = aes(x = bas24_nhd_pdvlot)) + 
  geom_bar() + theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS 24 Vac Lot Sentiment") + facet_wrap(~bas24_dem_gender)
  

vac_bldg_sex24 <- ggplot(data = bas24, mapping = aes(x = bas24_nhd_pdvbldg)) + 
  geom_bar() + theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS 24 Vac Bldg Sentiment") + facet_wrap(~bas24_dem_gender)

vac_bldg_sex25 <- ggplot(data = bas25, mapping = aes(x = bas25_nhd_pbvbld)) + 
  geom_bar() + theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  labs(x="BAS 25 Vac Bldg Sentiment") + facet_wrap(~bas25_dem_gender)
```

## Association Between Vacant Buildings and other Neighborhood Indicators from BAS Data

```{r}
#| label: VBN and Other Neighborhood Indicators

bas24_vac_blg_vs_walksafe <- bas24 %>% 
  select(bas24_nhd_pdvbldg, bas24_nhd_walksafe) %>% 
  group_by(bas24_nhd_pdvbldg) %>% 
  count(bas24_nhd_walksafe)

# Grouped bar plot
ggplot(data = bas24_vac_blg_vs_walksafe, 
       aes(x = bas24_nhd_pdvbldg, y = n, fill = bas24_nhd_walksafe)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  ylab("Frequency") + xlab("Vacant Building Sentiment") + 
  theme(axis.text.x = element_text(angle = 30, vjust = 0.8)) + 
  ggtitle("Bar chart of safe walking sentiment by vacant building sentiment")

# Mosaic plot - relative frequencies by vacant building sentiment
ggplot(data = bas24) +
  ggmosaic::geom_mosaic(aes(x = product(bas24_nhd_pdvbldg), 
                            fill = bas24_nhd_walksafe)) + 
  theme(axis.text.x = element_text(angle = 30, vjust = 0.8))
```
